---
title: "IST719 Final Project"
author: "Cory Vanderpoel"
date: "2026-02-15"
output: pdf_document
---

Patterns in Chicago Crime: A Multi‑Dimensional Exploration


About the data:
• Dataset description of two sentences.
This dataset contains individual crime incident records reported to the Chicago Police Department from January 2001 onward. Each row represents a single reported crime including its classification, location coordinates, whether an arrest was made, whether it was a domestic incident, and the police beat and district. The data supports crime prediction, hotspot analysis, temporal pattern detection, and public safety research.
Columns:
  id                       : Unique record identifier
  case_number              : Chicago Police Department case number
  date                     : Date and time of incident occurrence
  block                    : Partially redacted street address
  iucr                     : Illinois Uniform Crime Reporting code
  primary_type             : Primary crime classification (e.g., THEFT, BATTERY, ASSAULT)
  description              : Detailed crime description
  location_description     : Type of location (e.g., STREET, APARTMENT, SIDEWALK) #Explore this
  arrest                   : Whether an arrest was made (true/false)
  domestic                 : Whether the incident was domestic-related (true/false)
  beat                     : Police beat where incident occurred
  district                 : Police district number
  ward                     : City council ward number
  community_area           : Community area code
  year                     : Year of incident
  latitude                 : Latitude coordinate of incident
  longitude                : Longitude coordinate of incident
  updated_on               : Date record was last updated

• My compelling story of two sentences.
This poster examines spatial, temporal, and categorical patterns in Chicago crime to reveal how incidents cluster and evolve across the city. By integrating multiple dimensions of the data, the analysis highlights trends that are relevant for understanding and modeling urban risk. 

• My audience (who would care about it) of two sentences.
Insurance companies rely on accurate risk assessment to set premiums, evaluate claims exposure, and understand neighborhood‑level trends. Crime patterns—especially when examined across time, location, and type—offer valuable insight into potential losses and risk concentration. This motivates a deeper exploration of how different dimensions of Chicago crime interact.

Questions I plan to answer
- How do crime rates vary across Chicago neighborhoods when considering both crime type and time of year?
- Which types of crime are most strongly associated with specific geographic areas, and how might these patterns influence risk exposure?
- Is arrest likelihood related to the combination of crime type and location, and what might this imply for claims predictability?


```{r}
#Libraries
library(dplyr)
library(ggplot2)
library(lubridate)
library(hexbin)
library(ggmap)
library(RColorBrewer)
library(wordcloud)
library(tidytext)
library(tidyverse)
library(sf)
library(MetBrewer)
library(svglite)

#Standard Color Pallette for a clean poster
pal <- met.brewer("Hiroshige")
pal <- colorRampPalette(pal)(24)
```


Importing my data:
```{r}
crime.data <- read.csv("chicago crimes.csv"
                       , header = TRUE
                       , stringsAsFactors = FALSE
                       )

```

Getting a quick look:
```{r}
head(crime.data)
```

```{r}
#Data Cleanup
crime.data$DateObject <- mdy_hms(crime.data$Date) #from string to date object

crime_group_map <- c( #Re-mapping for my purposes
  #Violent Crime
  "BATTERY" = "Violent",
  "ASSAULT" = "Violent",
  "ROBBERY" = "Violent",
  "HOMICIDE" = "Violent",
  "CRIM SEXUAL ASSAULT" = "Violent",

  #Property Crime
  "THEFT" = "Property",
  "BURGLARY" = "Property",
  "MOTOR VEHICLE THEFT" = "Property",
  "ARSON" = "Property",
  "CRIMINAL DAMAGE" = "Property",

  #Drug / Vice
  "NARCOTICS" = "Drug/Vice",
  "PROSTITUTION" = "Drug/Vice",
  "GAMBLING" = "Drug/Vice",

  #Public Order
  "WEAPONS VIOLATION" = "Public Order",
  "INTERFERENCE WITH PUBLIC OFFICER" = "Public Order",
  "PUBLIC PEACE VIOLATION" = "Public Order",

  #White Collar / Fraud
  "DECEPTIVE PRACTICE" = "White Collar",
  "FORGERY" = "White Collar",
  "FRAUD" = "White Collar"
)

crime.data <- crime.data %>%
  mutate(
    Crime.Group = crime_group_map[Primary.Type],
    Crime.Group = if_else(is.na(Crime.Group), "Other", Crime.Group)
  )

crime.data <- crime.data %>%
  mutate(
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude)
  )


crime.data <- crime.data %>% #Mapping for my purposes - need to make this column less granular
  mutate(
    location.group = case_when(
      str_detect(Location.Description, regex("APARTMENT|RESIDENCE|HOME|YARD|PORCH|DRIVEWAY", ignore_case = TRUE)) ~ "Residential",
      str_detect(Location.Description, regex("STREET|SIDEWALK|ALLEY|PARKING LOT", ignore_case = TRUE)) ~ "Street/Sidewalk",
      str_detect(Location.Description, regex("STORE|GAS STATION|RESTAURANT|SHOP|MARKET", ignore_case = TRUE)) ~ "Commercial Retail",
      str_detect(Location.Description, regex("CTA|TRAIN|BUS|VEHICLE|TAXI|PLATFORM", ignore_case = TRUE)) ~ "Transit/Vehicle",
      str_detect(Location.Description, regex("SCHOOL|COLLEGE|DAYCARE|UNIVERSITY", ignore_case = TRUE)) ~ "School/Childcare",
      str_detect(Location.Description, regex("PARK|BEACH|PLAYGROUND|LAKEFRONT", ignore_case = TRUE)) ~ "Parks/Outdoor",
      str_detect(Location.Description, regex("GOVERNMENT|POLICE|HOSPITAL|COURT", ignore_case = TRUE)) ~ "Government/Institutional",
      str_detect(Location.Description, regex("FACTORY|WAREHOUSE|CONSTRUCTION", ignore_case = TRUE)) ~ "Industrial/Warehouse",
      str_detect(Location.Description, regex("BAR|CLUB|THEATER|ARENA", ignore_case = TRUE)) ~ "Entertainment",
      TRUE ~ "Other/Unknown"
    )
  )



head(crime.data)
```

Descriptive Plots:
```{r}
crimeDistribution <- crime.data %>%
  mutate(hour = hour(DateObject)) %>%
  ggplot(aes(x = hour)) +
  geom_histogram(binwidth = 1, fill = pal[3], color = "white") +
  scale_x_continuous(
    breaks = c(0, 6, 12, 18),
    labels = format(strptime(c(0, 6, 12, 18), "%H"), "%I %p")
  ) +
  scale_y_continuous(labels = NULL) +
  labs(
    title = "Distribution of Crime by Hour",
    x = "Hour of Day",
    y = "Total Crime Volume"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x = element_text(size = 10)
  )

ggsave(
  "PlotsForPoster/crimeDistribution.pdf",
  plot = last_plot(),
  width = 8,
  height = 6,
  device = cairo_pdf
)

crimeDistribution
```

```{r}
commonCrimes <- crime.data %>%
  count(Primary.Type) %>%
  slice_max(n, n = 15) %>%   #top 15 types
  ggplot(aes(x = reorder(Primary.Type, n), y = n)) +
  geom_col(fill = pal[8]) +
  coord_flip() +
  labs(
    title = "Most Common Crime Types",
    x = "Crime Type",
    y = "Total Volume"
  ) +
  theme_minimal(base_size = 14)

ggsave(
  "PlotsForPoster/commonCrimes.pdf",
  plot = last_plot(),
  width = 8,
  height = 6,
  device = cairo_pdf
)

commonCrimes
```

```{r}
monthlyCrimes <- crime.data %>%
  mutate(month = month(DateObject, label = TRUE)) %>%
  count(month) %>%
  ggplot(aes(x = month, y = n)) +
  geom_col(fill = pal[2]) +
  labs(
    title = "Crime Frequency by Month",
    x = "Month",
    y = "Number of Incidents"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

ggsave(
  "PlotsForPoster/monthlyCrimes.pdf",
  plot = last_plot(),
  width = 8,
  height = 6,
  device = cairo_pdf
)

monthlyCrimes
```

```{r}
arrestOutcomes <- crime.data %>%
  mutate(Arrest = if_else(Arrest == "true", "Yes", "No")) %>% #Remapping for readability
  count(Arrest) %>%
  ggplot(aes(x = Arrest, y = n, fill = Arrest)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = pal[c(1,4)]) +
  labs(
    title = "Arrest Outcomes",
    x = "Arrest Made",
    y = "Count"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

ggsave(
  "PlotsForPoster/arrestOutcomes.pdf",
  plot = last_plot(),
  width = 8,
  height = 6,
  device = cairo_pdf
)

arrestOutcomes
```


Heatmap
```{r}
seasonalPatterns <- crime.data %>%
  mutate(month = month(DateObject, label = TRUE)) %>%
  count(Crime.Group, month) %>%
  group_by(Crime.Group) %>%
  mutate(pct = n / sum(n)) %>%   #normalize within type
  ggplot(aes(x = month, y = fct_reorder(Crime.Group, pct, .fun = max), fill = pct)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "C") +
  labs(
    title = "Seasonal Patterns in Crime Types ",
    subtitle = "Normalized to highlight seasonal risk patterns",
    x = "Month",
    y = NULL,
    fill = "Percent"
  ) +
  theme_minimal(base_size = 14)

ggsave(
  "PlotsForPoster/seasonalPatterns.pdf",
  plot = seasonalPatterns,
  width = 8,
  height = 6,
  device = cairo_pdf
)

seasonalPatterns
```

Density Maps
```{r}
chicago_map <- st_read("Neighborhoods_2012b_20260330.geojson")
crime.clean <- crime.data %>%
  filter(
    between(Longitude, -87.94, -87.52),
    between(Latitude, 41.64, 42.02)
  )
crime.clean <- crime.clean %>%
  filter(Longitude != 0, Latitude != 0)

crime.recent <- crime.clean %>% filter(Year >= 2020)


```

Overall Crime Density
```{r}
register_stadiamaps(key = "331001f0-0400-4b42-bd5a-0d7317b5595f")

bbox <- c(
  left   = -87.82,
  bottom = 41.65,
  right  = -87.55,
  top    = 42.0
)

base <- get_stadiamap(
  bbox = bbox,
  zoom = 12,
  maptype = "stamen_toner_lite"
)

overallDenisty <- ggmap(base) +
  stat_density2d(
    data = crime.recent,
    aes(x = Longitude, y = Latitude, fill = after_stat(level), alpha = after_stat(level)),
    geom = "polygon",
    contour = TRUE
  ) +
  scale_fill_viridis_c(option = "inferno") +
  scale_alpha(range = c(0.2, 0.7), guide = "none") +
  labs(
    title = "Chicago Crime Density",
    subtitle = "Overall Activity"
  ) +
  theme_void()

ggsave(
  "PlotsForPoster/overallDenisty.pdf",
  plot = overallDenisty,
  width = 8,
  height = 6,
  device = cairo_pdf
)

overallDenisty
```

```{r}
arrestLikelyhood <- crime.data %>%
  count(District, Crime.Group, Arrest) %>%
  group_by(District, Crime.Group) %>%
  mutate(pct_arrest = n / sum(n)) %>%
  filter(Arrest == "true") %>%
  ggplot(aes(x = Crime.Group, y = pct_arrest, fill = factor(District))) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = pal) +
  labs(
    title = "Arrest Likelihood by Crime Type and District",
    x = "Crime Type",
    y = "Arrest Rate",
    fill = "District"
  ) +
  theme_minimal(base_size = 14)

ggsave(
  "PlotsForPoster/arrestLikelyhood.pdf",
  plot = arrestLikelyhood,
  width = 8,
  height = 6,
  device = cairo_pdf
)

arrestLikelyhood
```



Violent Crimes
```{r}
violent.crime <- crime.recent %>%
  filter(Crime.Group == "Violent")

base <- get_stadiamap(
  bbox = bbox,
  zoom = 12,
  maptype = "stamen_toner_lite"
)

violentCrimeMap <- ggmap(base) +
  stat_density2d(
    data = violent.crime,
    aes(x = Longitude, y = Latitude, fill = after_stat(level), alpha = after_stat(level)),
    geom = "polygon",
    contour = TRUE
  ) +
  scale_fill_viridis_c(option = "inferno") +
  scale_alpha(range = c(0.2, 0.7), guide = "none") +
  labs(
    title = "Violent Crimes",
    subtitle = " (2020-Present)"
  ) +
  theme_void() +
  theme(
    legend.position = "none",
    plot.margin = margin(0, 0, 0, 0)
  )

ggsave(
  "PlotsForPoster/violentCrimes.pdf",
  plot = violentCrimeMap,
  width = 8,
  height = 6,
  device = cairo_pdf
)

violentCrimeMap
```
Property Crime
```{r}
property.crime <- crime.recent %>%
  filter(Crime.Group == "Property")

base <- get_stadiamap(
  bbox = bbox,
  zoom = 12,
  maptype = "stamen_toner_lite"
)

propertyCrimes <- ggmap(base) +
  stat_density2d(
    data = property.crime,
    aes(x = Longitude, y = Latitude, fill = after_stat(level), alpha = after_stat(level)),
    geom = "polygon",
    contour = TRUE
  ) +
  scale_fill_viridis_c(option = "inferno") +
  scale_alpha(range = c(0.2, 0.7), guide = "none") +
  labs(
    title = "Property Crimes",
    subtitle = " (2020-Present)"
  ) +
  theme_void() +
  theme(
    legend.position = "none",
    plot.margin = margin(0, 0, 0, 0)
  )


ggsave(
  "PlotsForPoster/propertyCrimes.pdf",
  plot = propertyCrimes,
  width = 8,
  height = 6,
  device = cairo_pdf
)

propertyCrimes
```


Drug/Vice Crime
```{r}
vice.crimes <- crime.recent %>%
  filter(Crime.Group == "Drug/Vice")

base <- get_stadiamap(
  bbox = bbox,
  zoom = 12,
  maptype = "stamen_toner_lite"
)

viceCrimes <- ggmap(base) +
  stat_density2d(
    data = vice.crimes,
    aes(x = Longitude, y = Latitude, fill = after_stat(level), alpha = after_stat(level)),
    geom = "polygon",
    contour = TRUE
  ) +
  scale_fill_viridis_c(option = "inferno") +
  scale_alpha(range = c(0.2, 0.7), guide = "none") +
  labs(
    title = "Drug/Vice Crimes",
    subtitle = " (2020-Present)"
  ) +
  theme_void() +
  theme(
    legend.position = "none",
    plot.margin = margin(0, 0, 0, 0)
  )

ggsave(
  "PlotsForPoster/viceCrimes.pdf",
  plot = viceCrimes,
  width = 8,
  height = 6,
  device = cairo_pdf
)

viceCrimes
```

All Other
```{r}
other.crimes <- crime.recent %>%
  filter(!Crime.Group %in% c("Violent", "Property", "Drug/Vice"))

base <- get_stadiamap(
  bbox = bbox,
  zoom = 12,
  maptype = "stamen_toner_lite"
)

allOtherCrimes <- ggmap(base) +
  stat_density2d(
    data = other.crimes,
    aes(x = Longitude, y = Latitude, fill = after_stat(level), alpha = after_stat(level)),
    geom = "polygon",
    contour = TRUE
  ) +
  scale_fill_viridis_c(option = "inferno") +
  scale_alpha(range = c(0.2, 0.7), guide = "none") +
  labs(
    title = "All Other Crimes",
    subtitle = " (2020-Present)"
  ) +
  theme_void() +
  theme(
    legend.position = "none",
    plot.margin = margin(0, 0, 0, 0)
  )

ggsave(
  "PlotsForPoster/allOtherCrimes.pdf",
  plot = allOtherCrimes,
  width = 8,
  height = 6,
  device = cairo_pdf
)

allOtherCrimes
```


```{r}
arrest.rates <- crime.data %>%
  group_by(location.group, Crime.Group) %>%
  summarise(
    arrests = sum(Arrest == 'true', na.rm = TRUE),
    total = n(),
    arrest_rate = arrests / total,
    .groups = "drop"
  )

arrestGroupings <- ggplot(arrest.rates, aes(x = location.group, y = arrest_rate, fill = Crime.Group)) +
  geom_col(position = "stack", color = "white") +
  scale_fill_viridis_d(option = "C") +
  labs(
    title = "Arrest Rate by Location Category",
    x = "Location Category",
    y = "Arrest Rate",
    fill = "Crime Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  ) +
  coord_flip()

ggsave(
  "PlotsForPoster/arrestGroupings.pdf",
  plot = arrestGroupings,
  width = 8,
  height = 6,
  device = cairo_pdf
)

arrestGroupings
```


```{r}
library(tigris)
options(tigris_use_cache = TRUE)

#Chicago boundary
chi <- places(state = "IL", cb = TRUE) %>%
  filter(NAME == "Chicago") %>%
  st_transform(26971)  #Illinois East NAD83

#Neighborhoods (optional)
hoods <- neighborhoods <- tracts(state = "IL", county = "Cook", cb = TRUE) %>%
  st_transform(26971)

plotBackground <- ggplot() +
  geom_sf(data = chi, fill = "grey90", color = NA) +
  geom_sf(data = hoods, fill = NA, color = "grey70", size = 0.2, alpha = 0.4) +
  theme_void() +
  theme(
    plot.background = element_rect(fill = "white", color = NA)
  )
ggsave("PlotsForPoster/chicago_basemap.svg", plot = plotBackground, width = 24, height = 36, device = svglite)

```


Tangle visualization demo
```{r}
options(tigris_use_cache = TRUE)

#Chicago boundary
chi <- places(state = "IL", cb = TRUE) %>%
  filter(NAME == "Chicago") %>%
  st_transform(4326)


#Neighborhoods
hoods <- tracts(state = "IL", county = "Cook", cb = TRUE) %>%
  st_transform(4326)


crime <- crime.clean

#Ensure factors are consistent
crime$Crime.Group <- factor(crime$Crime.Group)
crime$location.group <- factor(crime$location.group)
crime$DateObject <- as.POSIXct(crime$Date,
                               format = "%m/%d/%Y %I:%M:%S %p",
                               tz = "America/Chicago")
crime$year <- year(crime$DateObject)

crime.distint <- crime %>%
  group_by(Crime.Group, location.group, year, Latitude,Longitude) %>%
  summarize(n_points = n(), .groups = "drop")

crime_dedup <- crime %>%
  group_by(Crime.Group, location.group, year) %>%
  distinct(Longitude, Latitude, .keep_all = TRUE) %>%
  ungroup()

output_dir <- "C:/Users/coryv/OneDrive/Documents/Master's Program - Syracuse University/07 - IST719/Homework/Final Project/TangleDemo/heatmaps"
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

bbox <- c(
  left   = -87.94011,
  bottom = 41.64454,
  right  = -87.52398,
  top    = 42.02304
)

#Get basemap (you can switch to "toner-lite" or others)
chicago_map <- get_stadiamap(
  bbox,
  zoom = 11,
  maptype = "stamen_toner_lite"
)

crime_groups <- levels(crime_dedup$Crime.Group)
location_groups <- levels(crime_dedup$location.group)
years <- sort(unique(crime_dedup$year))

for (cg in crime_groups) {
  for (lg in location_groups) {
    for (yr in years) {

      message("Rendering: ", cg, " / ", lg, " / ", yr)

      #Filter data
      df <- crime_dedup %>%
        filter(Crime.Group == cg,
               location.group == lg,
               year == yr)

      #Skip empty subsets
      if (nrow(df) <= 10) next

      p <- ggplot() +
        
        #Basemap from get_stadiamap()
        annotation_raster(
          chicago_map,
          xmin = bbox["left"],
          xmax = bbox["right"],
          ymin = bbox["bottom"],
          ymax = bbox["top"]
        ) +
        
        #Chicago boundary
        geom_sf(data = chi, fill = "grey90", color = NA) +
        
        #Neighborhood outlines
        geom_sf(data = hoods, fill = NA, color = "grey70", size = 0.2, alpha = 0.4) +
        
        #Density heatmap
        stat_density2d(
          data = df,
          aes(x = Longitude, y = Latitude, fill = ..level.., alpha = ..level..),
          geom = "polygon",
          contour = TRUE
        ) +
        
        #Coordinate system + bounding box
        coord_sf(
          crs = 4326,
          xlim = c(bbox["left"], bbox["right"]),
          ylim = c(bbox["bottom"], bbox["top"])
        ) +
        
        #Aesthetics
        scale_fill_viridis_c(option = "magma") +
        scale_alpha(range = c(0.2, 0.7), guide = FALSE) +
        theme_void() +
        theme(legend.position = "none")

      safe_cg <- gsub("/", "_", cg)
      safe_lg <- gsub("/", "_", lg)

      filename <- file.path(output_dir, paste0(safe_cg, "_", safe_lg, "_", yr, ".png"))
      
      ggsave(
        filename = filename,
        plot = p,
        width = 6,
        height = 4,
        dpi = 150
      )
    }
  }
}
```